---
title: "CCBR1155 CS031014 and CS029689 Joint Analysis"
author: "Samantha Sevilla"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Project info
This analysis report comparison two peak calling methods. [SEACR](https://github.com/FredHutch/SEACR) which includes the "normal.relaxed" peak threhold and [MACS2](https://hbctraining.github.io/Intro-to-ChIPseq/lessons/05_peak_calling_macs.html) which includes the "narrowPeak" threshold.

| CS ID | Type | Species | CCBR | Notes | 
| -- | -- | -- | -- | -- |
| CS031014 | Cut and Run | Hg38 | 1155 | this does not include INFB |
| CS029689 | Cut and Run | Hg38 | 1155 | this includes INFB |

## Links
- [CS031014](https://hpcnihapps.cit.nih.gov/genome/cgi-bin/hgTracks?db=hg38&lastVirtModeType=default&lastVirtModeExtraState=&virtModeType=default&virtMode=0&nonVirtPosition=&position=chrX%3A15560138%2D15602945&hgsid=138042_bQaCEQ730iKs2v633AtElgyIVxFO)

- [CS029689](https://hpcnihapps.cit.nih.gov/genome/cgi-bin/hgTracks?db=hg38&lastVirtModeType=default&lastVirtModeExtraState=&virtModeType=default&virtMode=0&nonVirtPosition=&position=chrX%3A15560138%2D15602945&hgsid=138044_F8trGjLAIsZ8fDb3lBslAIdLOr8E)

```{r paths_lists, echo=FALSE, warning=FALSE, include=FALSE}
############################################################
# project information
############################################################
#paths
parent_dir="~/../../Volumes/ccbr1155-1/"
input_car_dir1 = "~/../../Volumes/ccbr1155-1/CS031014/"
contrast_subpath1=paste0(input_car_dir1,"/carlisle_220920/results/peaks/contrasts/")

input_car_dir2 = "~/../../Volumes/ccbr1155-1/CS029689/"
contrast_subpath2=paste0(input_car_dir2,"/carlisle_220920/results/peaks/contrasts/")

output_dir = "~/../../Volumes/ccbr1155-1/CS031014_CS029689/r_analysis_220919/"
output_car_dir=output_dir
dir.create(file.path(output_dir), showWarnings = FALSE)
```

```{r user_params}
############################################################
# set global variables 
############################################################
padj_cutoff = 0.05
species = "Homo sapiens"
log2fc_cutoff_car = 1
minSize_gene_set=15
method="fragments" #AUC or fragments
dedup_status="dedup"
scalesfbymean="Y"
rawcountsprescaled="N"

#http://www.gsea-msigdb.org/gsea/msigdb/collections.jsp
db_list=c("C1","C2:BIOCARTA","C2:KEGG","C2:REACTOME","C2:WIKIPATHWAYS","C5:MF","C5:BP","C5:CC","H")

# group to ignore
ignore_groups=c()
```

```{r, include=FALSE}
## if there is a subset gene list, select files
gene_file = paste0(parent_dir,"docs/apm_alpha_genelist.csv")
pi_gene_df = read.csv(gene_file)

if (species=="Homo sapiens"){
  genome="hg38"
  txdb="TxDb.Hsapiens.UCSC.hg38.knownGene"
} else if (species=="Mus Musculus"){
  genome="mm10"
}
```

```{r echo=FALSE, warning=FALSE, include=FALSE}
# package list
list.of.packages=c("DESeq2","dplyr","ggplot2","ggrepel","ggpubr","tidyverse",
                   "ggVennDiagram",txdb,"karyoploteR","RColorBrewer","EnhancedVolcano",
                   "plotly","fgsea","msigdbr","gridExtra","kableExtra","clusterProfiler","UpSetR","ggupset",
                   "ggimage", "ChIPseeker","pander")
  
#install as needed
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) BiocManager::install(new.packages)

# source functions
source('CAR_functions.R')

# load packages
invisible(lapply(list.of.packages, library, character.only = TRUE))

#negate
`%ni%` <- Negate(`%in%`)

#anno
if (species=="Homo sapiens"){
  txdb<-TxDb.Hsapiens.UCSC.hg38.knownGene
  annodb="org.Hs.eg.db"
}
options(ChIPseeker.downstreamDistance = 0)
```

## **Data Processing**
```{r contrast, include=FALSE,warning=FALSE, message=FALSE}
####################
# run metadata input
####################
#groups input
groups_df1 = read.csv(paste0(input_car_dir1,"groups.tab"),sep="\t",header = FALSE)[,c("V1","V2")]
colnames(groups_df1) = c("sampleid","group")
rownames(groups_df1)=groups_df1$sampleid

groups_df2 = read.csv(paste0(input_car_dir2,"groups.tab"),sep="\t",header = FALSE)[,c("V1","V2")]
colnames(groups_df2) = c("sampleid","group")
rownames(groups_df2)=groups_df2$sampleid
```

## **Data Analysis**  {.tabset}
For each sample, all peaks are first annotated. Three summary graphics are then created:

1. Pie chart of all annotation terms of ALL peaks

2. Upset plot of all annotation terms of ALL peaks showing relationships of annotation terms

3. Bar chart of the distribution of transcriptin factor binding loci by feature size

Data was then filtered include only significant peaks. Three pie charts were then created:

1. Pie chart of all annotation terms of SIGNIFICANT peaks

2. Pie chart of all annotation terms INCREASED of SIGNIFICANT peaks

3. Pie chart of all annotation terms DECREASED of SIGNIFICANT peaks

Finally, two volcano plots were created.

### 53_H3K4me3_vs_HN6_H3K4me3 {.tabset}

#### SEACR {.tabset}
```{r wo_s, warning=FALSE, echo=FALSE, message=FALSE}
#sample level 
contrast_id=c("53_H3K4me3_vs_HN6_H3K4me3")
groups_df=groups_df1
car_subpath=contrast_subpath1

# peak tool
norm_type_cutandrun="norm.relaxed" # set analysis type (norm.relaxed=SEACR, narrowPeak=MACS2)
extensions=c(paste0("__",dedup_status,"__",norm_type_cutandrun,".bed"))
```

##### Summary
```{r wo_ss, warning=FALSE, echo=FALSE, message=FALSE}
#analysis
peak_anno=main_prep_qc_secondary(contrast_id,exclusion_list)
chipseeker_plots(peak_anno)
```

##### Pie Charts
```{r wo_sp, warning=FALSE, echo=FALSE, message=FALSE}
merged_sig_df=create_sig_df(contrast_id)
merged_sig_df$peaktype="SEACR"
promoter_df=subset(merged_sig_df,shortAnno=="Promoter")
collapsed_df=create_collapsed_df(merged_sig_df)
main_piecharts_from_collapsed_secondary(contrast_id)
```

##### Volcano
```{r wo_sv, warning=FALSE, echo=FALSE, message=FALSE}
generate_volcano_plots(contrast_id,gene_list_in = "ON")
DT::datatable(read.csv(paste0(output_car_dir,"volcano_data_",contrast_id,".csv")))
```

#### MACS2 {.tabset}
```{r wo_m, warning=FALSE, echo=FALSE, message=FALSE}
# peak tool
norm_type_cutandrun="narrowPeak" # set analysis type (norm.relaxed=SEACR, narrowPeak=MACS2)
extensions=c(paste0("__",dedup_status,"__",norm_type_cutandrun))
```

##### Summary
```{r wo_ms, warning=FALSE, echo=FALSE, message=FALSE}
#analysis
peak_anno=main_prep_qc_secondary(contrast_id,exclusion_list)
chipseeker_plots(peak_anno)
```

##### Pie Charts
```{r wo_mp, warning=FALSE, echo=FALSE, message=FALSE}
merged_sig_df=create_sig_df(contrast_id)
merged_sig_df$peaktype="MACS2"
promoter_df=rbind(promoter_df,
                  subset(merged_sig_df,shortAnno=="Promoter"))
collapsed_df=create_collapsed_df(merged_sig_df)
main_piecharts_from_collapsed_secondary(contrast_id)
```

##### Volcano
```{r wo_mv, warning=FALSE, echo=FALSE, message=FALSE}
generate_volcano_plots(contrast_id,gene_list_in = "ON")
DT::datatable(read.csv(paste0(output_car_dir,"volcano_data_",contrast_id,".csv")))
```

#### Overlap 
Reviewing the overlapping genes between promoters, and those that are unique to either
MACS2 or SECAR.
```{r o1, warning=FALSE, echo=FALSE, message=FALSE}
gene_seacr=unique(subset(promoter_df,peaktype=="SEACR")$SYMBOL)
gene_macs2=unique(subset(promoter_df,peaktype=="MACS2")$SYMBOL)

print(paste0("The total number of genes identified in SEACR: ",length(gene_seacr)))
print(paste0("The total number of genes identified in MACS2: ",length(gene_macs2)))

overlap_list=gene_seacr[gene_seacr %in% gene_macs2]
overlap_list=unique(append(gene_macs2[gene_macs2 %in% gene_seacr],overlap_list))
print(paste0("The genes in common (",length(overlap_list),")"))
print(overlap_list)

print(paste0("The genes in SEACR only (",length(gene_seacr[gene_seacr %ni% gene_macs2]),")"))
gene_seacr[gene_seacr %ni% gene_macs2]

print(paste0("The genes in MACS2 only (",length(gene_macs2[gene_macs2 %ni% gene_seacr]),")"))
gene_macs2[gene_macs2 %ni% gene_seacr]
```

### 53_H4K20m3_vs_HN6_H4K20me3 {.tabset}

#### SEACR {.tabset}
```{r wo_s, warning=FALSE, echo=FALSE, message=FALSE}
#sample level 
contrast_id=c("53_H4K20m3_vs_HN6_H4K20me3")
groups_df=groups_df1
car_subpath=contrast_subpath1

# peak tool
norm_type_cutandrun="norm.relaxed" # set analysis type (norm.relaxed=SEACR, narrowPeak=MACS2)
extensions=c(paste0("__",dedup_status,"__",norm_type_cutandrun,".bed"))
```

##### Summary
```{r wo_ss, warning=FALSE, echo=FALSE, message=FALSE}
#analysis
peak_anno=main_prep_qc_secondary(contrast_id,exclusion_list)
chipseeker_plots(peak_anno)
```

##### Pie Charts
```{r wo_sp, warning=FALSE, echo=FALSE, message=FALSE}
merged_sig_df=create_sig_df(contrast_id)
merged_sig_df$peaktype="SEACR"
promoter_df=subset(merged_sig_df,shortAnno=="Promoter")
collapsed_df=create_collapsed_df(merged_sig_df)
main_piecharts_from_collapsed_secondary(contrast_id)
```

##### Volcano
```{r wo_sv, warning=FALSE, echo=FALSE, message=FALSE}
generate_volcano_plots(contrast_id,gene_list_in = "ON")
DT::datatable(read.csv(paste0(output_car_dir,"volcano_data_",contrast_id,".csv")))
```

#### MACS2 {.tabset}
```{r wo_m, warning=FALSE, echo=FALSE, message=FALSE}
# peak tool
norm_type_cutandrun="narrowPeak" # set analysis type (norm.relaxed=SEACR, narrowPeak=MACS2)
extensions=c(paste0("__",dedup_status,"__",norm_type_cutandrun))
```

##### Summary
```{r wo_ms, warning=FALSE, echo=FALSE, message=FALSE}
#analysis
peak_anno=main_prep_qc_secondary(contrast_id,exclusion_list)
chipseeker_plots(peak_anno)
```

##### Pie Charts
```{r wo_mp, warning=FALSE, echo=FALSE, message=FALSE}
merged_sig_df=create_sig_df(contrast_id)
merged_sig_df$peaktype="MACS2"
promoter_df=rbind(promoter_df,
                  subset(merged_sig_df,shortAnno=="Promoter"))
collapsed_df=create_collapsed_df(merged_sig_df)
main_piecharts_from_collapsed_secondary(contrast_id)
```

##### Volcano
```{r wo_mv, warning=FALSE, echo=FALSE, message=FALSE}
generate_volcano_plots(contrast_id,gene_list_in = "ON")
DT::datatable(read.csv(paste0(output_car_dir,"volcano_data_",contrast_id,".csv")))
```

#### Overlap 
Reviewing the overlapping genes between promoters, and those that are unique to either
MACS2 or SECAR.
```{r o1, warning=FALSE, echo=FALSE, message=FALSE}
gene_seacr=unique(subset(promoter_df,peaktype=="SEACR")$SYMBOL)
gene_macs2=unique(subset(promoter_df,peaktype=="MACS2")$SYMBOL)

print(paste0("The total number of genes identified in SEACR: ",length(gene_seacr)))
print(paste0("The total number of genes identified in MACS2: ",length(gene_macs2)))

overlap_list=gene_seacr[gene_seacr %in% gene_macs2]
overlap_list=unique(append(gene_macs2[gene_macs2 %in% gene_seacr],overlap_list))
print(paste0("The genes in common (",length(overlap_list),")"))
print(overlap_list)

print(paste0("The genes in SEACR only (",length(gene_seacr[gene_seacr %ni% gene_macs2]),")"))
gene_seacr[gene_seacr %ni% gene_macs2]

print(paste0("The genes in MACS2 only (",length(gene_macs2[gene_macs2 %ni% gene_seacr]),")"))
gene_macs2[gene_macs2 %ni% gene_seacr]
```

### 5-3_H3K4me3_IFNb_vs_HN6_H3K4me3_IFNb {.tabset}

#### SEACR {.tabset}

```{r w_s, warning=FALSE, echo=FALSE, message=FALSE}
#sample level 
contrast_id=c("5-3_H3K4me3_IFNb_vs_HN6_H3K4me3_IFNb")
groups_df=groups_df2
car_subpath=contrast_subpath2

# peak tool
norm_type_cutandrun="norm.relaxed" # set analysis type (norm.relaxed=SEACR, narrowPeak=MACS2)
extensions=c(paste0("__",dedup_status,"__",norm_type_cutandrun,".bed"))
```

##### Summary
```{r w_ss, warning=FALSE, echo=FALSE, message=FALSE}
#analysis
peak_anno=main_prep_qc_secondary(contrast_id,exclusion_list)
chipseeker_plots(peak_anno)
```

##### Pie Charts
```{r w_sp, warning=FALSE, echo=FALSE, message=FALSE}
merged_sig_df=create_sig_df(contrast_id)
merged_sig_df$peaktype="SEACR"
promoter_df=subset(merged_sig_df,shortAnno=="Promoter")
collapsed_df=create_collapsed_df(merged_sig_df)
main_piecharts_from_collapsed_secondary(contrast_id)
```

##### Volcano
```{r w_sv, warning=FALSE, echo=FALSE, message=FALSE}
generate_volcano_plots(contrast_id,gene_list_in = "ON")
DT::datatable(read.csv(paste0(output_car_dir,"volcano_data_",contrast_id,".csv")))
```

#### MACS2 {.tabset}
```{r w_m, warning=FALSE, echo=FALSE, message=FALSE}
# peak tool
norm_type_cutandrun="narrowPeak" # set analysis type (norm.relaxed=SEACR, narrowPeak=MACS2)
extensions=c(paste0("__",dedup_status,"__",norm_type_cutandrun))
```

##### Summary
```{r w_ms, warning=FALSE, echo=FALSE, message=FALSE}
#analysis
peak_anno=main_prep_qc_secondary(contrast_id,exclusion_list)
chipseeker_plots(peak_anno)
```

##### Pie Charts
```{r w_mp, warning=FALSE, echo=FALSE, message=FALSE}
merged_sig_df=create_sig_df(contrast_id)
merged_sig_df$peaktype="MACS2"
promoter_df=rbind(promoter_df,
                  subset(merged_sig_df,shortAnno=="Promoter"))
collapsed_df=create_collapsed_df(merged_sig_df)
main_piecharts_from_collapsed_secondary(contrast_id)
```

##### Volcano
```{r w_mv, warning=FALSE, echo=FALSE, message=FALSE}
generate_volcano_plots(contrast_id,gene_list_in = "ON")
DT::datatable(read.csv(paste0(output_car_dir,"volcano_data_",contrast_id,".csv")))
```

#### Overlap 
Reviewing the overlapping genes between promoters, and those that are unique to either
MACS2 or SECAR.
```{r o2, warning=FALSE, echo=FALSE, message=FALSE}
gene_seacr=unique(subset(promoter_df,peaktype=="SEACR")$SYMBOL)
gene_macs2=unique(subset(promoter_df,peaktype=="MACS2")$SYMBOL)

print(paste0("The total number of genes identified in SEACR: ",length(gene_seacr)))
print(paste0("The total number of genes identified in MACS2: ",length(gene_macs2)))

overlap_list=gene_seacr[gene_seacr %in% gene_macs2]
overlap_list=unique(append(gene_macs2[gene_macs2 %in% gene_seacr],overlap_list))
print(paste0("The genes in common (",length(overlap_list),")"))
print(overlap_list)

print(paste0("The genes in SEACR only (",length(gene_seacr[gene_seacr %ni% gene_macs2]),")"))
gene_seacr[gene_seacr %ni% gene_macs2]

print(paste0("The genes in MACS2 only (",length(gene_macs2[gene_macs2 %ni% gene_seacr]),")"))
gene_macs2[gene_macs2 %ni% gene_seacr]
```