---
title: "CCBR1155 CS031014_and_CS028891 Joint Analysis"
author: "Samantha Sevilla"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Project info
This analysis report comparison two projects

| CS ID | Type | Species | CCBR | Notes | 
| -- | -- | -- | -- | -- |
| CS031014 | Cut and Run | Hg38 | 1155 |  |
| CS028891 | RNA Seq | Hg38 | 1155 |  |

## Links
- [CS031014](https://hpcnihapps.cit.nih.gov/genome/cgi-bin/hgTracks?db=hg38&lastVirtModeType=default&lastVirtModeExtraState=&virtModeType=default&virtMode=0&nonVirtPosition=&position=chr9%3A107479012%2D107495610&hgsid=138535_dtOi8FAAAvU6lnDCSPlrAnn80sBm  )

- [CS028891](https://hpcnihapps.cit.nih.gov/genome/cgi-bin/hgTracks?db=hg38&lastVirtModeType=default&lastVirtModeExtraState=&virtModeType=default&virtMode=0&nonVirtPosition=&position=chr12%3A123657495%2D123707504&hgsid=138556_akkFeKcMAIw5J1o0CnHd1XZ4FrAM)

```{r paths_lists, echo=FALSE, warning=FALSE, include=FALSE}
############################################################
# project information
############################################################
#paths
parent_dir="~/../../Volumes/CUTRUN/"
input_car_dir = "~/../../Volumes/CUTRUN/analysis/CS031014/"
car_subpath=paste0(input_car_dir,"/carlisle_221206/results/peaks/contrasts/")
bed_subpath=paste0(input_car_dir,"/carlisle_221206/results/peaks/gopeaks/")
output_car_dir=paste0(input_car_dir,"r_analysis_230118/")

input_rna_dir= "~/../../Volumes/CUTRUN/analysis/CS028891/"
output_rna_dir=paste0(input_rna_dir,"r_analysis_221223/")

output_dir = "~/../../Volumes/CUTRUN/analysis/CS031014_CS028891_complete/r_analysis_230118/"
img_dir = paste0(output_dir,"img/")
dir.create(file.path(output_dir), showWarnings = FALSE)
dir.create(file.path(img_dir), showWarnings = FALSE)
```

```{r user_params}
############################################################
# set global variables 
############################################################
padj_cutoff = 0.05
species = "Homo sapiens"
log2fc_cutoff_car=0.3785 #FC of 1.3
log2fc_cutoff_rna=0.3785 #FC of 1.3
minSize_gene_set=15
cs_id="CS031014"

dedup_status="dedup"
method="fragments"
neg_control="igG_1"
contrast_list=c("Promoter","Intron","Exon","5'UTR","3'UTR")

#http://www.gsea-msigdb.org/gsea/msigdb/collections.jsp
db_list=c("C1","C2:BIOCARTA","C2:KEGG","C2:REACTOME","C2:WIKIPATHWAYS","C5:MF","C5:BP","C5:CC","H")

# group to ignore
ignore_groups=c()

## pull cell_cycle genes
gene_file = paste0(parent_dir,"docs/gene_list_cellcycle_gsea.csv")
gene_df_cellcycle = read.csv(gene_file)

# pull cell cycle genes
gene_file = paste0(parent_dir,"docs/gene_list_emt.csv")
gene_df_emt = read.csv(gene_file)

# dbs
if (species=="Homo sapiens"){
  genome="hg38"
  txdb="TxDb.Hsapiens.UCSC.hg38"
} else if (species=="Mus Musculus"){
  genome="mm10"
}
```

```{r echo=FALSE, warning=FALSE, include=FALSE}
# package list
list.of.packages=c("DESeq2","dplyr","ggplot2","ggrepel","ggpubr","tidyverse",
                   "ggVennDiagram","karyoploteR","RColorBrewer","EnhancedVolcano",
                   "plotly","fgsea","msigdbr","gridExtra","kableExtra","clusterProfiler",
                   "pheatmap","grid","gplots","pander","ChIPseeker","tidyr","TxDb.Mmusculus.UCSC.mm10.knownGene",
                   "org.Mm.eg.db", "TxDb.Hsapiens.UCSC.hg19.knownGene","org.Hs.eg.db", "TxDb.Hsapiens.UCSC.hg38.knownGene")
  
#install as needed
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) BiocManager::install(new.packages)

# source functions
source(paste0(parent_dir,'github/CAR_core_functions.R'))

# load packages
invisible(lapply(list.of.packages, library, character.only = TRUE))

#negate
`%ni%` <- Negate(`%in%`)
```

## Annotation information
The genomic region of the peak is reported in annotation column using [ChIPseeker](http://bioconductor.org/packages/devel/bioc/vignettes/ChIPseeker/inst/doc/ChIPseeker.html). Since some annotation may overlap, the following priority listing was set:

- Promoter

- 5’ UTR

- 3’ UTR

- Exon

- Intron

- Distal (defined as "distal intergenic")

## Questions  {.tabset}

### Question 1
*1) What is the genome-wide distribution pattern of H3K4me3 in HN-6 cells before and after siSMYD3 treatment (pie chart)?*
```{r q1, warning=FALSE, echo=FALSE}
# run function
contrast_id=c("53_H3K4me3_vs_HN6_H3K4me3")
peak_type="narrowGo"
single_sample_picharts(contrast_id,peak_type,neg_control)
main_piecharts_from_collapsed_nonsig(contrast_id)
main_piecharts_from_collapsed_secondary(contrast_id)
```

### Question 2
*2) List of genes corresponding to differential H3K4me3 (log2FC by 30%) present on promoters or gene bodies (include only promoters, introns and exons).*
```{r q2, warning=FALSE, echo=FALSE}
contrast_id="53_H3K4me3_vs_HN6_H3K4me3"
main_differential_genes(contrast_id,contrast_list)
```

### Question 3
*3) Heatmap of the above list of genes (increased peaks, decreased peaks).*
```{r q3, warning=FALSE, echo=FALSE}
contrast_id="53_H3K4me3_vs_HN6_H3K4me3"
fshort="promoter_intron_exon_3UTR_5UTR"
gene_bodies_list=c("Promoter","Intron","Exon","5'UTR","3'UTR")
generate_heat_map_differential(contrast_id,n_up=5,n_down=47,fshort) #5; 45
```

### Question 4
- A gene list was created of the genes significant in either CUT&RUN (CAR) or the RNASeq experiment (pvalue and log2foldchange).  Venn diagrams were created, showing the overlap and uniquely significant genes for each experiment.

- Genes that were significant in the CAR experiment (IE left side of venn diagram) were plotted on a Karoyplot to show gene placement. Color mapping indiciates up/down regulation in the CAR only experiment or in the CAR and RNA experiments.

- The genes that were found to be significantly differentiated in both datasets are then presented in a datatable.

- Next, RNASeq data was filtered to only include genes, regardless of significance, in the CAR data. This allowed for the peakAnnotations (IE Intron, Exon) to be added to each gene. The process was then repeated, with a list of genes significant in either experiment being created after filtering for the annotation (IE "Promoter","Intron","Exon") of interest. If applicable, Venn diagrams, karyoplots, and datatables were created.

*4A) Of the decreasing H3K4me3 peaks annotated to promoters/gene bodies of genes, what is the overlap with genes differentially downregulated (by 30% or more) on the RNA-seq database?*
```{r q4down, warning=FALSE, echo=FALSE}
contrast_id_car="53_H3K4me3_vs_HN6_H3K4me3"
contrast_id_rna="CRISPR_53_without_IFNb-parental_HN6_without_IFNb"

# run genebodies, down
gene_bodies_list=c("Promoter","Intron","Exon","5'UTR","3'UTR")
fshort=paste0("genebodies_down")
main_differential_overlap(subset_list=gene_bodies_list,
                          rna_regulation="down",
                          car_regulation="down")
```

*4B) Of the increasing H3K4me3 peaks annotated to promoters/gene bodies of genes, what is the overlap with genes differentially downregulated (by 30% or more) on the RNA-seq database?*

```{r q4up, warning=FALSE, echo=FALSE}
contrast_id_car="53_H3K4me3_vs_HN6_H3K4me3"
contrast_id_rna="CRISPR_53_without_IFNb-parental_HN6_without_IFNb"

##run all,up
#subset_list=c("all")
#fshort=paste0("all_up")
#main_differential_overlap(subset_list=subset_list,
#                          rna_regulation="up",
#                          car_regulation="up")

# run genebodies, up
gene_bodies_list=c("Promoter","Intron","Exon","5'UTR","3'UTR")
fshort=paste0("genebodies_up")
main_differential_overlap(subset_list=gene_bodies_list,
                          rna_regulation="up",
                          car_regulation="up")
```

*4C) Create heatmap of the overlapping genes described in 4A and 4B*
```{r q4hm, warning=FALSE, echo=FALSE}
contrast_id_car="53_H3K4me3_vs_HN6_H3K4me3"
contrast_id_rna="CRISPR_53_without_IFNb-parental_HN6_without_IFNb"
main_differential_overlap_all(contrast_id_car,contrast_id_rna)
```


### Question 5
*5) Overlap with up- or downregulated cell cycle genes sorted from question (4).*
```{r}
contrast_id_car="53_H3K4me3_vs_HN6_H3K4me3"
contrast_id_rna="CRISPR_53_without_IFNb-parental_HN6_without_IFNb"
gene_bodies_list=c("Promoter","Intron","Exon","5'UTR","3'UTR")
contrast_id=paste0(contrast_id_car,"_",contrast_id_rna)

gene_list=gene_df_cellcycle$Human
gene_list_name="cell_cycle_genes"
main_piecharts_from_genelist(contrast_id,gene_list,gene_list_name)
```


### Question 6
*6) Overlap with up- or downregulated EMT genes sorted from question (4).*
```{r}
contrast_id_car="53_H3K4me3_vs_HN6_H3K4me3"
contrast_id_rna="CRISPR_53_without_IFNb-parental_HN6_without_IFNb"
gene_bodies_list=c("Promoter","Intron","Exon","5'UTR","3'UTR")
contrast_id=paste0(contrast_id_car,"_",contrast_id_rna)

gene_list=gene_df_emt$Human
gene_list_name="emt_genes"
main_piecharts_from_genelist(contrast_id,gene_list,gene_list_name)
```

### Question 7
*7) Generate H3K4me3 CUT&RUN volcano plot (significant peaks affected by 30%) and a separate RNA-seq volcano plot, and mark cell cycle genes in each.*

```{r q7, warning=FALSE, echo=FALSE, message=FALSE}
gene_list=gene_df_cellcycle$Human
gene_list_name="Cell_Cycle_genes"

contrast_id=c("53_H3K4me3_vs_HN6_H3K4me3")
generate_volcano_plots("CAR",contrast_id,gene_list_in="ON")

contrast_id="CRISPR_53_without_IFNb-parental_HN6_without_IFNb"
generate_volcano_plots("RNA",contrast_id,gene_list_in="ON")
```

### Question 8
*8) Generate H3K4me3 CUT&RUN volcano plot (significant peaks affected by 30%) and a separate RNA-seq volcano plot, and mark EMT genes in each.*
```{r q8, warning=FALSE, echo=FALSE, message=FALSE}
gene_list=gene_df_emt$Human
gene_list_name="EMT_genes"

contrast_id=c("53_H3K4me3_vs_HN6_H3K4me3")
generate_volcano_plots("CAR",contrast_id,gene_list_in="ON")

contrast_id="CRISPR_53_without_IFNb-parental_HN6_without_IFNb"
generate_volcano_plots("RNA",contrast_id,gene_list_in="ON")
```

### Question 9
*9) Generate heatmap ranked by CHIP seq intensity for all H3K4me3 peaks (Figure 5C)*

### Question 10
*10)Generate boxplots of ChIP seq intensity for H3K4me3 peaks within promoters or gene body regions.*
```{r}
contrast_id=c("53_H3K4me3_vs_HN6_H3K4me3")
#generate_intensity_boxplot(contrast_id)
```

### Question 11
*11) What is the genome-wide distribution pattern of H4K20me3 in HN-6 cells before and after SMYD3 knockout (pie chart)?*
```{r q11, warning=FALSE, echo=FALSE}
# run function
contrast_id=c("53_H4K20m3_vs_HN6_H4K20me3")
main_piecharts_from_collapsed_nonsig(contrast_id)
main_piecharts_from_collapsed_secondary(contrast_id)
```

### Question 12
*12) List of genes corresponding to differential H4K20m3 (log2FC by 30%) present on promoters or gene bodies (include only promoters, introns and exons).*
```{r q12, warning=FALSE, echo=FALSE}
contrast_id=c("53_H4K20m3_vs_HN6_H4K20me3")
contrast_list=c("Promoter","Intron","Exon","5'UTR","3'UTR")
main_differential_genes(contrast_id,contrast_list)
```

### Question 13
*13) Heatmap of the above list of genes (increased peaks, decreased peaks).*
```{r q13, warning=FALSE, echo=FALSE}
contrast_id=c("53_H4K20m3_vs_HN6_H4K20me3")
fshort="promoter_intron_exon_3UTR_5UTR"
gene_bodies_list=c("Promoter","Intron","Exon","5'UTR","3'UTR")
generate_heat_map_differential(contrast_id,n_up=49,n_down=50,fshort) #28up; 22down
```

### Question 14
- A gene list was created of the genes significant in either CUT&RUN (CAR) or the RNASeq experiment (pvalue and log2foldchange).  Venn diagrams were created, showing the overlap and uniquely significant genes for each experiment.

- Genes that were significant in the CAR experiment (IE left side of venn diagram) were plotted on a Karoyplot to show gene placement. Color mapping indiciates up/down regulation in the CAR only experiment or in the CAR and RNA experiments.

- The genes that were found to be significantly differentiated in both datasets are then presented in a datatable.

- Next, RNASeq data was filtered to only include genes, regardless of significance, in the CAR data. This allowed for the peakAnnotations (IE Intron, Exon) to be added to each gene. The process was then repeated, with a list of genes significant in either experiment being created after filtering for the annotation (IE "Promoter","Intron","Exon") of interest. If applicable, Venn diagrams, karyoplots, and datatables were created.
*14A) Of the decreasing H4K20me3 peaks annotated to promoters/gene bodies of genes, what is the overlap with genes differentially upregulated (by 30% or more) on the RNA-seq database?*
```{r q14down, warning=FALSE, echo=FALSE}
contrast_id_car="53_H4K20m3_vs_HN6_H4K20me3"
contrast_id_rna="CRISPR_53_without_IFNb-parental_HN6_without_IFNb"

#run all, down
# subset_list=c("all")
# fshort=paste0("all_down")
# main_differential_overlap(subset_list=subset_list,
#                           rna_regulation="up",
#                           car_regulation="down")

# run genebodies, down
gene_bodies_list=c("Promoter","Intron","Exon","5'UTR","3'UTR")
fshort=paste0("genebodies_down")
main_differential_overlap(subset_list=gene_bodies_list,
                          rna_regulation="up",
                          car_regulation="down") 
```

*14B) Of the increasing H3K4me3 peaks annotated to promoters/gene bodies of genes, what is the overlap with genes differentially downregulated (by 30% or more) on the RNA-seq database?*
```{r q14up, warning=FALSE, echo=FALSE}
contrast_id_car="53_H4K20m3_vs_HN6_H4K20me3"
contrast_id_rna="CRISPR_53_without_IFNb-parental_HN6_without_IFNb"

#run all, down
subset_list=c("all")
fshort=paste0("all_up")
# main_differential_overlap(subset_list=subset_list,
#                           rna_regulation="down",
#                           car_regulation="up")

# run genebodies, down
gene_bodies_list=c("Promoter","Intron","Exon","5'UTR","3'UTR")
fshort=paste0("genebodies_up")
main_differential_overlap(subset_list=gene_bodies_list,
                          rna_regulation="down",
                          car_regulation="up")
```


*14C) Create heatmap of the overlapping genes described in 4A and 4B*
```{r q14hm, warning=FALSE, echo=FALSE}
contrast_id_car="53_H4K20m3_vs_HN6_H4K20me3"
contrast_id_rna="CRISPR_53_without_IFNb-parental_HN6_without_IFNb"
main_differential_overlap_all(contrast_id_car,contrast_id_rna)
```

### Question 15
*15)cell cycle gene list: overlap with up- or downregulated genes sorted from question (4).*
```{r q15bcell, warning=FALSE, echo=FALSE, message=FALSE}
contrast_id_car=c("53_H4K20m3_vs_HN6_H4K20me3")
contrast_id_rna="CRISPR_53_without_IFNb-parental_HN6_without_IFNb"
gene_bodies_list=c("Promoter","Intron","Exon","5'UTR","3'UTR")
contrast_id=paste0(contrast_id_car,"_",contrast_id_rna)

gene_list=gene_df_cellcycle$Human
gene_list_name="cell_cycle_genes"
main_piecharts_from_genelist(contrast_id,gene_list,gene_list_name)
```

### Question 16
*16) EMT gene list: overlap with up- or downregulated genes sorted from question (4).*
```{r q16, warning=FALSE, echo=FALSE, message=FALSE}
contrast_id_car=c("53_H4K20m3_vs_HN6_H4K20me3")
contrast_id_rna="CRISPR_53_without_IFNb-parental_HN6_without_IFNb"
gene_bodies_list=c("Promoter","Intron","Exon","5'UTR","3'UTR")
contrast_id=paste0(contrast_id_car,"_",contrast_id_rna)

gene_list=gene_df_emt$Human
gene_list_name="emt_genes"
main_piecharts_from_genelist(contrast_id,gene_list,gene_list_name)
```

### Question 17
*17) Generate H3K4me3 CUT&RUN volcano plot (significant peaks affected by 30%) and a separate RNA-seq volcano plot, and mark cell cycle genes in each.*
```{r q17, warning=FALSE, echo=FALSE, message=FALSE}
gene_list=gene_df_cellcycle$Human
gene_list_name="Cell_Cycle_genes"

contrast_id=c("53_H4K20m3_vs_HN6_H4K20me3")
generate_volcano_plots("CAR",contrast_id,gene_list_in="ON")

contrast_id="CRISPR_53_without_IFNb-parental_HN6_without_IFNb"
generate_volcano_plots("RNA",contrast_id,gene_list_in="ON")
```

### Question 18
*18) Generate H3K4me3 CUT&RUN volcano plot (significant peaks affected by 30%) and a separate RNA-seq volcano plot, and mark EMT genes in each.*
```{r q18, warning=FALSE, echo=FALSE, message=FALSE}
gene_list=gene_df_emt$Human
gene_list_name="EMT_genes"

contrast_id=c("53_H4K20m3_vs_HN6_H4K20me3")
generate_volcano_plots("CAR",contrast_id,gene_list_in="ON")

contrast_id="CRISPR_53_without_IFNb-parental_HN6_without_IFNb"
generate_volcano_plots("RNA",contrast_id,gene_list_in="ON")
```

### Question 19
*19) Generate heatmap ranked by CHIP seq intensity for all H3K4me3 peaks (Figure 5C)*

### Question 20
*20) Generate boxplots of ChIP seq intensity for H4K20 peaks within promoters or gene body regions.*
```{r}
contrast_id=c("53_H4K20m3_vs_HN6_H4K20me3")
#generate_intensity_boxplot(contrast_id)
```